---
format: 
   revealjs:
    embed-resources: true
    theme: serif
    slide-level: 3
    slide-number: true
    show-slide-number: all
    preview-links: auto
    number-sections: true
    link-color: orange
title: "Design declaration, diagnosis, and redesign "
subtitle: "MIDA and more"
author: "Graeme Blair, Alex Coppock, Macartan Humphreys"
bibliography: bib.bib
---


```{r, include=FALSE}
source("setup.R")
run <- FALSE
```



# `DeclareDesign`: A deeper dive {#secdd2}

## Steps in an experimental design

We start with a simple experimental design with all four elements of MIDA and then show ways to extend.

* Variations to *M* and *I* are supported by the `fabricatr` package (and others)
* Variations to *D*  are supported by the `randomizr` package (and others)
* Variations to *A*  are supported by the `estimatr` package (and others)


### A simple experimental design {.smaller}


```{r}

N <- 100
b <- .5

design <- 
  declare_model(N = N, U = rnorm(N), 
                potential_outcomes(Y ~ b * Z + U)) + 
  declare_assignment(Z = simple_ra(N), Y = reveal_outcomes(Y ~ Z)) + 
  declare_inquiry(ate = mean(Y_Z_1 - Y_Z_0)) + 
  declare_estimator(Y ~ Z, inquiry = "ate", .method = lm_robust)

```

New elements:

* `declare_model` can be used much like `mutate` with multiple columns created in sequence
* the `potential_outcomes` function is a special function that creates potential outcome columns for different values of `Z`
* when you assign a treatment that affects an outcome you can use `reveal_outcome` to reveal the outcome; `Z` and `Y` are  default names
* when you declare an estimator you should normally include a label, specify an inquiry, and provide the method to be used (`lm_robust` is default)

### Steps: Order matters

e.g. If you sample before defining the inquiry you get a different inquiry to if you sample after you define the inquiry


```{r}
design_1 <- 
  declare_model(N = 1000, X = rep(0:1, N/2), Y = X + rnorm(N)) + 
  declare_sampling(S= strata_rs(strata = X, strata_prob = c(.2, .8))) +
  declare_inquiry(m = mean(Y))

design_1 |> draw_estimands()
```

### Steps: Order matters

e.g. If you sample before defining the inquiry you get a different inquiry to if you sample after you define the inquiry

```{r}
design_2 <- 
  declare_model(N = 1000, X = rep(0:1, N/2), Y = X + rnorm(N)) + 
  declare_inquiry(m = mean(Y)) +
  declare_sampling(S= strata_rs(strata = X, strata_prob = c(.2, .8))) 

design_2 |> draw_estimands()
```

## M: Key extensions to model declaration

### Hierarchical data 

You can generate hierarchical data like this:

```{r}
M <- 
  declare_model(
    households = add_level(
      N = 100, 
      N_members = sample(c(1, 2, 3, 4), N, 
                         prob = c(0.2, 0.3, 0.25, 0.25), replace = TRUE)
    ),
    individuals = add_level(
      N = N_members, 
      age = sample(18:90, N, replace = TRUE)
    )
  )
```

### Hierarchical data 

You can generate hierarchical data like this:

```{r}
M() |> head() |> kable(digits = 2) |> kable_styling(font_size = 20)
```

### Panel data   {.smaller}

You can generate panel data like this:

```{r}
M <- 
  declare_model(
    countries = add_level(
      N = 196, 
      country_shock = rnorm(N)
    ),
    years = add_level(
      N = 100, 
      time_trend = 1:N,
      year_shock = runif(N, 1, 10), 
      nest = FALSE
    ),
    observation = cross_levels(
      by = join_using(countries, years),
      observation_shock = rnorm(N),
      Y = 0.01 * time_trend + country_shock + year_shock + observation_shock 
    )
  )
```

### Panel data

You can generate panel data like this:

```{r}
M() |> head() |> kable(digits = 2) |> kable_styling(font_size = 20)
```


### Preexisting data

```{r, eval = FALSE}
M <- 
  declare_model(
    data = baseline_data,
    attitudes = sample(1:5, N, replace = TRUE)
  )
```

###  Steps

You can repeat steps and play with the order, always conscious of the direction of the pipe

```{r}
design <- 
  declare_model(N = N, X = rep(0:1, N/2)) +
  declare_model(U = rnorm(N), potential_outcomes(Y ~ b * Z * X + U)) + 
  declare_assignment(Z = block_ra(blocks = X), Y = reveal_outcomes(Y ~ Z)) + 
  declare_inquiry(ate = mean(Y_Z_1 - Y_Z_0)) + 
  declare_inquiry(cate = mean(Y_Z_1[X==0] - Y_Z_0[X==0])) + 
  declare_estimator(Y ~ Z, inquiry = "ate", label = "ols") + 
  declare_estimator(Y ~ Z*X, inquiry = "cate", label = "fe")

```


###  You can generate multiple columns together

```{r multiplecolumns}

M2 <-
  declare_model(
    draw_multivariate(c(X1, X2) ~ MASS::mvrnorm(
      n = 1000,
      mu = c(0, 0),
      Sigma = matrix(c(1, 0.3, 0.3, 1), nrow = 2)
    )))

```

###  You can generate multiple columns together

```{r}
M2() |> head() |> kable(digits = 2) |> kable_styling(font_size = 28) 
```

###  Cluster structures with cluster correlations

```{r}
M <-
  declare_model(households = add_level(N = 1000),
                individuals = add_level(
                  N = 4,
                  X = draw_normal_icc(
                    mean = 0,
                    clusters = households,
                    ICC = 0.65
                  )
                ))
```

###  Cluster structures with cluster correlations

```{r}
model <- lm_robust(X ~ households, data = M())
model$adj.r.squared
```


## I: Inquiries

### Inquiries using predefined potential outcomes

Many causal inquiries are simple summaries of potential outcomes:

::: {style="font-size: 20px;"}
| Inquiry                                                    | Units                                  | Code                                                                 |  
| ---------------------------------------------------------- | -------------------------------------- |--------------------------------------------------------------------- |
| Average treatment effect in a finite population (PATE)     | Units in the population                | `mean(Y_D_1 - Y_D_0)`                                                | 
| Conditional average treatment effect (CATE) for X = 1      | Units for whom X = 1                   | `mean(Y_D_1[X == 1] - Y_D_0[X == 1])`                                | 
| Complier average causal effect (CACE)                      | Complier units                         | `mean(Y_D_1[D_Z_1 > D_Z_0] - Y_D_0[D_Z_1 > D_Z_0])`                  | 
| Causal interactions of $D_1$ and $D_2$                     | Units in the population                | `mean((Y_D1_1_D2_1 - Y_D1_0_D2_1) - (Y_D1_1_D2_0 - Y_D1_0_D2_0))`    | 
:::

Generating potential outcomes columns gets you far

### Inquiries using functions

Often though we need to define inquiries as a function of continuous variables. For this generating a potential outcomes function can make life easier. This helps for:

* Continuous quantities
* Spillover quantities
* Complex counterfactuals

### Complex counterfactuals

Here is an example of using functions to define complex counterfactuals:

```{r}

f_M <- function(X, UM) 1*(UM < X)
f_Y <- function(X, M, UY) X + M - .4*X*M + UY

design <- 
  declare_model(N = 100,
                X = simple_rs(N),
                UM = runif(N),
                UY = rnorm(N),
                M = f_M(X, UM),
                Y = f_Y(X, M, UY)) +
  declare_inquiry(Q1 = mean(f_Y(1, f_M(0, UM), UY) - f_Y(0, f_M(0, UM), UY)))

design |> draw_estimands() |> kable() |> kable_styling(font_size = 20)
```


### Complex counterfactuals {.smaller}

Here is an example of using functions to define effects of continuous treatments.

```{r}

f_Y <- function(X, UY) X - .25*X^2 + UY

design <- 
  declare_model(N = 100,
                X  = rnorm(N),
                UY = rnorm(N),
                Y = f_Y(X, UY)) +
  declare_inquiry(
    Q1 = mean(f_Y(X+1, UY) - f_Y(X, UY)),
    Q2 = mean(f_Y(1, UY) - f_Y(0, UY)),
    Q3 = (lm_robust(Y ~ X)|> tidy())[2,2]
    )

design |> draw_estimands() |> kable() |> kable_styling(font_size = 20)
```

which one is the ATE?

## D 

###  Assignment schemes

The `randomizr` package has a set of functions for different types of block and cluster assignments.


* Simple random assignment: "Coin flip" or Bernoulli random assignment. All units have the same probability of assignment:  `simple_ra(N = 100, prob = 0.25)`
* Complete random assignment: Exactly m of N units are assigned to treatment, and all units have the same probability of assignment m/N `complete_ra(N = 100, m = 40)`

###  Assignment schemes

* Block random assignment: Complete random assignment within pre-defined blocks. Units within the same block have the same probability of assignment $m_b / N_b$: `block_ra(blocks = regions)`
* Cluster random assignment:  Whole groups of units are assigned to the same treatment condition. `cluster_ra(clusters = households)`                         * Block-and-cluster assignment: Cluster random assignment within blocks of clusters `block_and_cluster_ra(blocks = regions, clusters = villages)`                               

###  Assignment schemes

You can combine these in various ways. For examples with saturation random assignment first clusters are assigned to a saturation level, then units within clusters are assigned to treatment conditions according to the saturation level:

```{r, eval = FALSE}
saturation = cluster_ra(clusters = villages, 
                        conditions = c(0, 0.25, 0.5, 0.75))

block_ra(blocks = villages, prob_unit = saturation)
```

## A 

### A: Answers: terms {.smaller}

By default `declare_estimates()` assumes you are interested in the *first term after the constant* from the output of an estimation procedure.

But you can say what you are interested in directly using `term` and you can also associate different terms with different quantities of interest using `inquiry`.

```{r terms}
design <-
  declare_model(
    N = 100,
    X1 = rnorm(N),
    X2 = rnorm(N),
    X3 = rnorm(N),
    Y = X1 - X2 + X3 + rnorm(N)
  ) +
  declare_inquiries(ate_2 = -1, ate_3 = 1) +
  declare_estimator(Y ~ X1 + X2 + X3,
                    term = c("X2", "X3"),
                    inquiry = c("ate_2", "ate_3"))

design  |> run_design()  |> kable(digits = 2) |> kable_styling(font_size = 20)
```

### A: Answers: terms  {.smaller}

Sometimes it can be confusing what the names of a term is but you can figure this by running the estimation strategy directly. Here's an example where the names of a term might be confusing.

```{r}
lm_robust(Y ~ A*B, 
          data = data.frame(A = rep(c("a",  "b"), 3), 
                            B = rep(c("p", "q"), each = 3), 
                            Y = rnorm(6))) |>
  coef() |> kable() |> kable_styling(font_size = 20)
```

The names as they appear in the output here is the name of the term that `declare_estimator` will look for.


### A: Answers: other packages  {.smaller}

`DeclareDesign` works natively with `estimatr` but you you can use whatever packages you like. You do have to make sure though that estimatr gets as input a nice tidy dataframe of estimates, and that might require some tidying.

```{r glm}
design <- 
  declare_model(N = 1000, U = runif(N), 
                potential_outcomes(Y ~ as.numeric(U < .5 + Z/3))) + 
  declare_assignment(Z = simple_ra(N), Y = reveal_outcomes(Y ~ Z)) + 
  declare_inquiry(ate = mean(Y_Z_1 - Y_Z_0)) + 
  declare_estimator(Y ~ Z, inquiry = "ate", 
                    .method = glm, 
                    family = binomial(link = "probit"))
```

Note that we passed additional arguments to `glm`; that's easy. 

It's not a good design though. Just look at the diagnosis:

### A: Answers: other packages

```{r, eval = FALSE}
diagnose_design(design)
```

```{r}
if(run)
  diagnose_design(design) |> write_rds("saved/probit.rds")

read_rds("saved/probit.rds") |> 
  reshape_diagnosis() |>
  kable() |> 
  kable_styling(font_size = 20)
```

Why is it so terrible?


### A: Answers: other packages {.smaller}

Because the probit estimate does not target the ATE directly; you need to do more work to get there. 

You essentially have to write a function to get the estimates, calculate  the quantity of interest and other stats, and turn these into a nice dataframe.

Luckily you can use the `margins` package with  `tidy` to create a `.summary` function which you can pass to `declare_estimator` to do all this for you

```{r margins}

tidy_margins <- function(x) 
  broom::tidy(margins::margins(x, data = x$data), conf.int = TRUE)

design <- design +  
  declare_estimator(Y ~ Z, inquiry = "ate", 
                    .method = glm, 
                    family = binomial(link = "probit"),
                    .summary = tidy_margins,
                    label = "margins")
```

### A: Answers: other packages

```{r}
if(run)
  diagnose_design(design) |> write_rds("saved/probit_2.rds")

read_rds("saved/probit_2.rds") |> reshape_diagnosis() |> kable() |> 
  kable_styling(font_size = 20)
```


Much better


## Advanced diagnosis

### diagnosands code

```{r, eval = FALSE}
DeclareDesign:::default_diagnosands
```

```{r, eval = FALSE}
    mean_estimand <- mean(estimand)
    mean_estimate <- mean(estimate)
    bias <- mean(estimate - estimand)
    sd_estimate <- sd(estimate)
    rmse <- sqrt(mean((estimate - estimand)^2))
    power <- mean(p.value <= alpha)
    coverage <- mean(estimand <= conf.high & estimand >= conf.low)
```


### More diagnosands

```{r, eval = FALSE}

    mean_se = mean(std.error)
    type_s_rate = mean((sign(estimate) != sign(estimand))[p.value <= alpha])
    exaggeration_ratio = mean((estimate/estimand)[p.value <= alpha])
    var_estimate = pop.var(estimate)
    mean_var_hat = mean(std.error^2)
    prop_pos_sig = estimate > 0 & p.value <= alpha
    mean_ci_length = mean(conf.high - conf.low)

```

### Custom diagnosands

```{r, warning = FALSE}

my_diagnosands <-
  declare_diagnosands(median_bias = median(estimate - estimand))

diagnose_design(simplest_design, diagnosands = my_diagnosands, sims = 10) |>
  reshape_diagnosis() |> kable() |> kable_styling(font_size = 20)
```

### Adding diagnosands to a design

```{r, warning = FALSE}
simplest_design <- 
  set_diagnosands(simplest_design, my_diagnosands)

simplest_design |> diagnose_design(sims = 10)|>
  reshape_diagnosis() |> kable() |> kable_styling(font_size = 20)

```

### Diagnosing in groups {.smaller}


You can partition the simulations data frame into groups before calculating diagnosands. 

```{r, eval = FALSE}
grouped_diagnosis <- 
  
  simplest_design |>
  diagnose_design(
    make_groups = vars(significant = p.value <= 0.05),
    sims = 500
  )
```


```{r, echo = FALSE}
if(run)
  simplest_design |>
  
  
  set_diagnosands() |>

    diagnose_design(
    make_groups = vars(significant = p.value <= 0.05),
    sims = 500
  ) |>
  write_rds("saved/group_rep.rds")

grouped_diagnosis <-   read_rds("saved/group_rep.rds") 

grouped_diagnosis |>
  reshape_diagnosis() |>
  select(-Inquiry, - Estimator, -Outcome, -Term) |>
  kable(digits = 2) |> 
  kable_styling(font_size = 20)

```

Note especially the mean estimate,  the power, the coverage, the RMSE, and the bias.
(Bias is not large because we have both under and over estimates)

### Significance filter

```{r, fig.height = 2, fig.width = 6}
grouped_diagnosis$simulations_df |>
  ggplot(aes(estimate, p.value, color = significant)) + geom_point()
```

### Multistage simulation 

* Usually a design simulation simulates "from the top": going from the beginning to the end of the design in each run and repeating
* But sometimes you might want to follow a tree like structure and simulate different steps a different number of times

### Multistage simulation illustration

Consider for instance this sampling design:

```{r}
sampling_design <- 
  
  declare_model(N = 500, Y = 1 + rnorm(N, sd = 10)) +
  declare_inquiry(Q = mean(Y)) +
  declare_sampling(S = complete_rs(N = N, n = 100)) + 
  declare_estimator(Y ~ 1)

```


### Multistage simulation illustration

Compare these two diagnoses:

```{r, warning = FALSE, eval = FALSE}
diagnosis_1 <- diagnose_design(sampling_design, sims = c(5000, 1, 1, 1)) 
diagnosis_2 <- diagnose_design(sampling_design, sims = c(1, 5000, 1, 1))
```



```{r, warning = FALSE, echo = FALSE}
if(run){
set.seed(1)
diagnose_design(sampling_design, sims = c(5000, 1, 1, 1)) |> write_rds("saved/multistage1.rds") 
diagnose_design(sampling_design, sims = c(1, 5000, 1, 1)) |> write_rds("saved/multistage2.rds")
}

list(diagnosis_1 = read_rds("saved/multistage1.rds") |> reshape_diagnosis(),
     diagnosis_2 = read_rds("saved/multistage2.rds") |> reshape_diagnosis()) |>
  bind_rows(.id = "diagnosis") |>
  select(-Design, -Inquiry, -Estimator, -Outcome, -Term) |>
  kable(digits = 2) |> 
  kable_styling(font_size = 20)
     

```

In the second the estimate is drawn just once. The SD of the estimate is lower. But the RMSE is not very different. 

### Illustration with tailored estimator and diagnosand

```{r}
my_estimator <- function(data) 
  data.frame(outcome = "Y", estimate = mean(data$Y), std.error = 0)

design <-
  declare_model(N = 5000, Y = rnorm(N)) +
  declare_inquiry(Y_bar_pop = mean(Y, 1)) +
  declare_sampling(S = complete_rs(N = N, n = 500)) +
  declare_inquiry(Y_bar_sample = mean(Y)) +
  declare_estimator(Y ~ 1, inquiry = "Y_bar_pop",label = "ols") +
  declare_estimator(handler = label_estimator(my_estimator),
                    inquiry = "Y_bar_sample",
                    label = "mean")
my_diagnosands <-
  declare_diagnosands(
    bias = mean(estimate - estimand),
    rmse = mean((estimate - estimand)^2)^.5,
    mean_se = mean(std.error))
```

### Diagnosis

```{r, eval = FALSE}
diagnose_design(design, diagnosands = my_diagnosands)
```


```{r, echo = FALSE}
if(run)
  diagnose_design(design, diagnosands = my_diagnosands) |> write_rds("saved/popsamp.rds")

read_rds("saved/popsamp.rds") |> reshape_diagnosis() |>   kable(digits = 2) |> kable_styling(font_size = 20)
```


## Advanced Redesign

### Redesign with vector arguments
When redesigning with arguments that are vectors,
use `list()` in redesign, with each list item representing a design you wish to create

```{r, eval = FALSE}

prob_each <- c(.1, .5, .4)

design_multi  <- 
  declare_model(N = 10) +
  declare_assignment(Z = complete_ra(N = N, prob_each = prob_each))

## returns two designs

designs <- design_multi |> 
  redesign(prob_each = list(c(.2, .5, .3), c(0, .5, .5)))
  
designs |> lapply(draw_data)
```


### Redesign warnings


A parameter has to be called correctly. And you get no warning if you misname.

```{r}
simplest_design_N  |> redesign(n = 200) |> draw_data() |> nrow()
```

why not 200?

### Redesign warnings


A parameter has to be called explicitly

```{r}
N <- 100

my_N <- function(n = N) n

simplest_design_N2 <- 
  
  declare_model(N = my_N(), Y = rnorm(N)) +
  declare_inquiry(Q = 0) +
  declare_estimator(Y ~ 1)

simplest_design_N2 |> redesign(N = 200) |> draw_data() |> nrow()
```

why not 200?

### Redesign warnings


A parameter has to be called explicitly

```{r}
N <- 100

my_N <- function(n = N) n

simplest_design_N2 <- 
  
  declare_model(N = my_N(N), Y = rnorm(N)) +
  declare_inquiry(Q = 0) +
  declare_estimator(Y ~ 1)

simplest_design_N2 |> redesign(N = 200) |> draw_data() |> nrow()
```


OK

### Redesign with a function

Here is an example of redesigning where the "parameter" is a function

```{r}
new_N <- function(n, factor = 1.31) n*factor

simplest_design_N2 |> redesign(my_N = new_N) |> draw_data() |> nrow()

```

